---
title: "Data Science for the Natural Environment (FA21)"
subtitle: "BI449 Data Science for the Natural Environment"
author: "Shannon J. O'Leary"
date: "`r Sys.Date()`"
knit: "bookdown::preview_chapter"
output:
  msmbstyle::msmb_html_book:
    highlight: tango
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    margin_references: FALSE
    css: msmb.css
bibliography: labmanual.bib
link-citations: yes
---

```{r include=FALSE}

library(msmbstyle)

# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      cache.extra = packageVersion('msmbstyle'),
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

options(htmltools.dir.version = FALSE)

source("scr/ggplot.R")

```

# Climate Change

**Learning Objectives**

After completing this tutorial you should be able to

* analyze earth's global temperatures to determine if they are increasing
* analyze CO<sub>2</sub> data to determine if atmospheric levels are increasing
* correlate CO<sub>2</sub> and global temperature to determine relationship
* compare current trends of with rates of change during pre-historic periods using ice core data
* interpret results to understand current climate change

Download the [03_ClimateChange](https://drive.google.com/drive/folders/100Zbhu8l_euEA9lyKWXtcfMhhQpn8vWb?usp=sharing) project folder. Once you have downloaded it, unzip the project directory into your `BI449` directory. You can open the `Rproj` for this module either by double clicking on it which will launch `Rstudio` or by opening `Rstudio` and then using `File > Open Project` or by clicking on the `Rproject` icon in the top right of your program window and selecting `Open Project`. 

Once you have opened a project you should see the project name in the top right corner^[Pro tip: If you run into issues where an Rmarkdown won't render or file paths aren't working (especially if things were working previously) one of your first steps should be to double check that the correct `Rproj` is loaded.].

You should have a `ClimChange.Rmd` in your project directory. Use that file to work through this set of activities - you will hand in your rendered ("knitted") rmarkdown file as your homework assignment.  

You will use this `Rmarkdown` file to record your answers. So, per usual, first thing in the `YAML` header, change the author to your name. Remember to use comments to describe your code; at minimum you should have one comment per code set^[You should do this even if you are adding code yourself or if you are using code from our manual, even if it isn't commented in the manual ... you may of course add as many comments as you need to be able to recall what you did]. Now that you have gained additional `Rmarkdown` skills use those to format your "report" that you will hand in. Code chunks should have comments, add answers to questions in plain text (not as code in chunks), use headers intentionally, and of course you have options like bullet points etc. as well.

Global change biology/ecology is an emerging field of research that is focused how changes in temperature and precipitation resulting from anthropogenic climate change will in turn alter the distribution, abundance, behavior, and physiology of on individual species, populations, communities, and ecosystems. In recent years, modeling has become increasingly sophisticated. Additionally, long-term and large-scale data sets are becoming increasingly available. This allows us to not only investigate impacts that can be observed now but also predict how different scenarios of global change will impact the biosphere. Typical lines of investigation include range shifts and invasions, how communities can be altered and the formation of new ones, impacts of urbanization, wildfires, and other natural catastrophes, loss of biodiversity, and changes in phenology along with understanding mechanisms of species/ecosystem resilience, and the interplay of biosphere, atmosphere, and hydrosphere. Many problems have both a temporal and spatial aspect making them especially complex and difficult to model. 

Climate regimes are defined by patterns of temperature and precipitation. Global temperature is a function of how much energy the earth receives and stores which in turn is influenced by three major factors:

1. The amount of energy received from the sun.
2. Reflection of energy by earth's surface.
3. Atmospheric composition (greenhouse gas effect).

During this unit we will start by taking a look at past and present rates of change of global temperature and CO<sub>2</sub> concentrations. Then we will take a look at some additional data sets to sharpen our visualization skills. 

Let's start by loading the packages we will need for this tutorial.

```{r}

# load libraries
library(knitr)             # kable() to make pretty tables
library(tidyverse)         # includes readr, tidyr, and dplyr to wrangle data & ggplot to plot
library(ggthemes)          # additional options for formatting plots
library(plotly)            # makes interactive plots possible
library(patchwork)         # allows us to combine plots

# turn of sci notation
options(scipen=999)

```


## Current rates of air temperature change

Let's start by taking a look at changes in global mean air temperatures.

`r msmbstyle::question_begin(label = "ques:cch-1")`

Sketch out what the patterns of air temperature over time would look like if the (A) earth is warming, (B) cooling, (C) not changing at all and discuss with your class mates (you do not need to include this in your Rmarkdown file). Determine what variables you would plot on the x-axis, y-axis and what the slope would look like.

`r msmbstyle::question_end()`

The air temperature data we will be using is compiled by the Goddard Institute for Space Studies (NASA) and can be accessed in [on their webpage](https://data.giss.nasa.gov/gistemp/.) which also describes how their data set was compiled and processed. Scroll down the page 
to the section `Tables of Global and Hemispheric Monthly Means and Zonal Annual Means` and download the [`CSV`](https://data.giss.nasa.gov/gistemp/tabledata_v4/GLB.Ts+dSST.csv) version of the `Global-mean monthly seasonal, and annual means.

`r msmbstyle::question(label = "ques:cch-2", text = "This data is presented as temperature anomalies, i.e. as deviations from the corresponding 1951 - 1980 mean. Explain what this means and argue why this is a good way to present this data..")`

After you have downloaded the data set place it in your `data` folder in your project directory and read it into your `Global Environment`.

`r msmbstyle::question(label = "ques:cch-3", text = "This is a csv formatted file, how can we read it into R?")`

`r msmbstyle::solution_begin()`

Correct! You need to use the `delim` argument. Let's try to read in the file.

```{r}

# read csv file
temperature <- read_delim("data/GLB.Ts+dSST.csv", delim = ",")

```

`r msmbstyle::solution_end()`

Before we move on, let's check our `temperature` object in the `Global Environment` to see if the file read in okay - spoiler alert, it did not.

**Troubleshooting Skills**: Your file didn't read in correctly. Let's figure out why. 

`r msmbstyle::question(label = "ques:cch-4", text = "What ideas do you have for us to track down the issue?")`


`r msmbstyle::solution_begin()`

A good starting point is to open the file in a text editor to see if we missed anything. Sure enough there seems to be an additional line at the beginning which is probably causing the issue. One way to fix this is to simply delete the extra line ... but if we want to have a reproducible workflow we want to avoid manually editing our data sets. Instead use `?read_delim` to pull up the help file for the function. You should find an argument called `skip` which will let us tell the function how many extra lines there are at the beginning of the file.

```{r}

# read csv file
temperature <- read_delim("data/GLB.Ts+dSST.csv", delim = ",", skip = 1)

head(temperature)

```

That seems to have done the trick.

`r msmbstyle::solution_end()`

Let's take a slightly more detailed look at our data set to make sure there aren't any additional changes we need to make, for example we need to determine if there are `NA` values and if they are properly formatted and we need to make sure all the columns are `numeric`.

`r msmbstyle::question(label = "ques:cch-5", text = "Take a look at your data frame, make sure that you know what information each column contains, identify if anything is out of order, and figure out how to clean up any issues.")`

`r msmbstyle::solution_begin()`

`NA` values have been specified as `***`, which has caused some columns to be formatted as `character` instead of `numeric`. We can use `replace()` to search every column (we specify this using `.` instead of a specific column name) and `mutate_if()` which tells R to check every column and if it is a character data type (`is.character`) to convert it to numeric (`as.numeric`).

```{r}

temperature <- temperature %>%
  replace(. == "***", NA) %>%
  mutate_if(is.character, as.numeric)

```

`r msmbstyle::solution_end()`

`r msmbstyle::question(label = "ques:cch-6", text = "Now that we are all set let's create a plot with the yearly mean global temperature anomaly across time. Conceptually, what parameters do we need to plot on the x and y-axis to accomplish this?")`

`r msmbstyle::solution_begin()`

We are going to use `ggplot2` which is part of the `tidyverse` to plot our figure. This will introduce you to some of the standard syntax ^[Fear not, we will work through the details of the framework on which `ggplot2` relies later on].

```{r eval=FALSE}

ggplot(temperature, aes(x = Year, y = `J-D`)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(color = "red", size = 1.5, se = FALSE) +
  labs(x = "year", y = "temperature anomaly [C]",
       title = "Annual mean global surface temperature relative to 1951-1980 average."
       caption = "source: NASA Goddard Institute for Space Studies")

```


```{r caption="Annual mean global surface temperature relative to 1951-1980 average.", echo=FALSE}

ggplot(temperature, aes(x = Year, y = `J-D`)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(color = "red", size = 1.5, se = FALSE) +
  labs(x = "year", y = "temperature anomaly [C]",
       title = "Annual mean global surface temperature relative to 1951-1980 average.",
       caption = "source: NASA Goddard Institute for Space Studies") +
  theme_standard

```

`r msmbstyle::solution_end()`

`r msmbstyle::question(label = "ques:cch-7", text = "Interpret the figure: Describe the change in the deviation of global temperature to the 1950-1980 mean. Explain what it means when values are negative or positive.")`

This figure gives us a qualitative view of changing global temperatures. To quantitatively assess the rate of change, we can fit a linear regression as `y = b + mx`. With `m` as the slope and `b` as the intercept to determine the rate of change. 

For this example you would define the rate of change as the change in temperature divided by the change in time for a certain time period. A more general definition would be that you are calculating the **slope** of the line you have fit through the data as the change-in-y divided by the change-in-x^[If you compare the two figures you should see that fitting a linear regression is an oversimplification but it will allow us to make a quantitative comparison].

```{r eval=FALSE}

# define formula for regression
my.formula <- y ~ x

ggplot(temperature, aes(x = Year, y = as.numeric(`J-D`))) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", size = 1, se = FALSE) +
  labs(x = "year", y = "temperature anomaly [C]",
       title = "Annual mean global surface temperature relative to 1951-1980 average.",
       caption = "source: NASA Goddard Institute for Space Studies")

```

```{r caption="Annual mean global surface temperature relative to 1951-1980 average."}

# define formula for regression
my.formula <- y ~ x

ggplot(temperature, aes(x = Year, y = as.numeric(`J-D`))) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", size = 1, se = FALSE) +
  labs(x = "year", y = "temperature anomaly [C]",
       title = "Annual mean global surface temperature relative to 1951-1980 average.",
       caption = "source: NASA Goddard Institute for Space Studies") +
  theme_standard

```

We can fit the linear regression using the function `lm()`.

```{r}

# fit linear regression
score_model <- lm(`J-D` ~ Year, data = temperature)

# view summary of results
summary(score_model)

```

`r msmbstyle::question_begin(label = "ques:cch-8")`

We are interested in the equation of the regression line, how well the data fits the line, and whether or not the regression is significant. We can use the output of the linear regression to determine that. The estimate columns shows the values for the intercept `b` and the slope `m` for the variable (in this case year).

Determine what the equation for our linear regression looks like and use this to determine the rate of change, including what the units would be.

`r msmbstyle::question_end()`


`r msmbstyle::question_begin(label = "ques:cch-9")`

The adjusted R2 value describes the proportion of variance of the dependent value explained by the independent variable. In our figure what is the dependent and the independent value? How well does the data fit our regression? 

`r msmbstyle::question_end()`


`r msmbstyle::question_begin(label = "ques:cch-10")`

The column `PR(>|t|)` is the probability of the coefficient for the variable. Is our regression significant?


`r msmbstyle::question_end()`


`r msmbstyle::question(label = "ques:cch-11", text = "Interpret the results of your analysis to determine whether you have evidence of a warming earth.")`


`r msmbstyle::question_begin(label = "ques:cch-12")`

Now you! 

1. Identify the time period where you observe the steepest rate of increase.
2. Create a subset of the data set to contain only that time frame. 
3. Plot you subset of data with a linear regression. 
4. Perform a statistical analysis to determine the rate of change.
5. Summarize your results and compare them to your results above.

`r msmbstyle::question_end()`


## Rate of change of atmospheric CO<sub>2</sub>

Next, let's determine the rate of change of atmospheric CO<sub>2</sub>. 

Dr Charles David Keeling (1928 - 2005) began collecting atmospheric CO<sub>2</sub> concentration data at the Mauna Observatory (Hawaii); this data set comprises the longest measurement of atmospheric CO<sub>2</sub> concentrations^[Longest measurement using intruments - we will see later that we can use proxy data to indirectly measure CO<sub>2</sub> levels for much longer time periods]. This data set has been fundamental to our understand the role of human activities such as fossil fuel burning in driving climate change.

`r msmbstyle::question(label = "ques:cch-13", text = "Briefly argue why Hawaii is a good location for this.")`

We can access the data set directly from the [Global Monitoring Laboratory](https://gml.noaa.gov/ccgg/trends/). Select the `Data` tab, then download the `csv` data set `Mauna Loa CO2 annual mean data`.

Once you have downloaded the data set, make sure to move it to the `data` folder of your project directory then read it into Rstudio.

```{r}

CO2 <- read_delim("data/co2_annmean_mlo.csv", delim = ",", skip = 55)

head(CO2)

```
`r msmbstyle::question(label = "ques:cch-14", text = "Explore your data set to determine if it is correctly formatted and briefly describe what information it contains - bullet points are fine.")`

`r msmbstyle::question(label = "ques:cch-15", text = "Plot the change in CO2 over time and use a linear regression to determine the rate of change in atmospheric CO2 over the entire data set based on your analysis. Argue how confident you are in these results.")`

```{r caption = "Change in annual mean global surface temperature relative to 1951-1980 average and annual mean atmospheric CO<sub>2</sub> concentrations.", fig.width=10, fig.height=5, echo=FALSE}

p1 <- temperature %>%
  filter(Year >= 1959) %>%
  ggplot(aes(x = Year, y = `J-D`)) +
  geom_line(color = "blue", size = 1.5) +
  geom_smooth(color = "red", size = 1, se = FALSE) +
  labs(x = "year", y = "temperature anomaly [C]",
       caption = "source: NASA Goddard Institute for Space Studies") +
  theme_standard

p2 <- ggplot(CO2, aes(x = year, y = mean)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(color = "red", size = 1.5, se = FALSE) +
  labs(x = "year", y = "mean atmospheric CO<sub>2</sub> [ppm]",
       caption = "source: Global Monitoring Laboratry/NOAA") +
  theme_standard

p1 + p2

```

`r msmbstyle::question(label = "ques:cch-16", text = "Compare your results of rates of change of average global temperature and atmospheric CO2 and explain the phenomenon that can be used to explain this.")`


## Comparison of current and pre-historic rates of change 

The most recent IPCC assessment has labled the increase in temperatures driving contemporary climate change as "unprecedented". Temperature have changed in earth's past, and temperatures have even been higher than what we are experiencing now. What is unprecedented is the rate at which this is occurring. Shall we investigate?

To do this we will need to look at past climate change. The two data sets we just took a look at are measurements of temperature and CO<sub>2</sub> using instrumentation, i.e. we have directly measured values for the parameters we are interested in at different points in time. Dr. Keeling was one of the first scientists to consider the importance of long-term monitoring sites. The rapid changes taking place in our environment have created a push to generate long-term data sets with a focus on making the accessible. We will take a look at some of these data sets later in this semester and you will likely use at least one of them for your own data science project.

Of course, we frequently have questions that might extend beyond data sets like the two we used above - How do we access data from before we had instrumentation and long-term measurement data sets set up? One way to do this we have to use so-called [proxy data sets](https://www.ncdc.noaa.gov/news/what-are-proxy-data).

`r msmbstyle::question(label = "ques:cch-17", text = "In general, a proxy is an intermidary or subsistute, i.e. it is a parameter that can be used to represent the value of something in a calculation. Paleoclimatologist use preserved physical characteristics of the environment to stand in for direct measurements using instruments, typical examples would be ice cores, tree rings, ocean sediments, or fossil pollen. Briefly discuss the pros and cons of proxy data.")`

For a deep dive you can check out [a detailed interactive brief on proxy data](https://interactive.carbonbrief.org/how-proxy-data-reveals-climate-of-earths-distant-past/).

Hundreds of ice cores extracted from polar ice have proven valuable to understanding changes in atmospheric chemistry over pre-historic time. Here, we can make use of the fact that as the ice is formed, air bubbles are trapped. Because these air bubbles have remained frozen, they still have the same composition of gases as at the time they were trapped. The depth of an ice core is correlated to time, deeper ice is older. In other words, ice cores form an archive of atmospheric conditions over time. We can directly measure CO<sub>2</sub> from the air bubbles trapped in the ice and we can measure isotopic ratios of oxygen in the water moleculesif  the core to derive temperature.

Vostok Ice core data set has been constructed using ice cores from the Vostock research station in the Antarctica and can be access through the [Carbon Dioxide Information Analysis Center](https://data.ess-dive.lbl.gov/portals/CDIAC).

Let's start by taking a look at the temperature data - download the data and load it into R^[We are using a slightly different method from before which allows us to directly download the data into our data folder. We are also using `read_table2()` from the `readr` package due to the fact that our text file is formatted using neither white space nor tabs.].

```{r}

# define url
url <- "http://cdiac.ess-dive.lbl.gov/ftp/trends/temp/vostok/vostok.1999.temp.dat"

# download file
download.file(url, "data/vostok_temp.txt")

# load dataset
vostok_temp <- read_table2("data/vostok_temp.txt",
                     skip = 60,
                     col_names = c("depth", "age_ice", "deuterium", "temperature"))

```

`r msmbstyle::question(label = "ques:cch-18", text = "Take a look at the data set and briefly describe what data is contained in the data set (you may want to take a peak at the original text file to get a better understanding).")`

Now we can create a plot of the temperature data over time. The age is recorded as years before present. For better visualization we will convert this to "thousand years ago" by dividing that number by 1,000.

```{r caption = "Temperature derived from Vostok ice core, Antarctica.", fig.height=6, fig.width=15}

ggplot(vostok_temp, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = 1) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)") +
  theme_standard

```

`r msmbstyle::question_begin(label = "ques:cch-19")`

Before we move at looking at rates of change, we first need to determine how to properly read this plot. Consider the following:

* What does it mean that time (on the x-axis) is represented as "Millenia before present", how does this differ from the other time-series we have plotted today?
* Temperature is being measured by proxy by looking at differences in isotope ratios; the data file lists this information as "Temperature variation". What does 0 C mean on this plot?
* Consider how long glacial and interglacial periods typically last - are we currently in a glacial or interglacial period?
* In what parts of the figure is temperature increasing/decreasing (consider slope)?

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:cch-20")`

Now add a trend line to our ice core temperature data and run a linear regression model. Argue whether or not you think this trend line is a good representation of long-term temperature change.

`r msmbstyle::question_end()`


`r msmbstyle::solution_begin()`

```{r}

ggplot(vostok_temp, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       title = "Temperature derived from Vostok ice core, Antarctica.",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)")

# fit linear regression
score_model <- lm(temperature ~ age_ice, data = vostok_temp)

# view summary of results
summary(score_model)

```


```{r caption = "Temperature derived from Vostok ice core, Antarctica.", fig.height=6, fig.width=10, eval=FALSE}

ggplot(vostok_temp, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       title = "Temperature derived from Vostok ice core, Antarctica.",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)") +
  theme_standard

```

`r msmbstyle::solution_end()`

In order to determine if the currently observed rate of change is "unprecedented" or not, let's identify past time periods with the fastest rate of change and calculate them.

To do this I will show you a little trick to plot an interactive figure.

```{r fig.height=6, fig.width=15}

p1 <- ggplot(vostok_temp, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = .75) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)")

ggplotly(p1)

```

Now you can use your cursor to identify individual points on the plot to select the subset you want plot.

```{r eval=FALSE}

# define time range
min_year_vostok <- 128357
max_year_vostok <- 138193

# filter data set + plot
vostok_temp_subset <- vostok_temp %>%
  filter(age_ice >= min_year_vostok & age_ice <= max_year_vostok)

ggplot(vostok_temp_subset, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       title = "Temperature change recorded in Vostok ice core (128357 - 138193 years ago).",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)")

```

```{r}

# define time range
min_year_vostok <- 128357
max_year_vostok <- 138193

# filter data set + plot
vostok_temp_subset <- vostok_temp %>%
  filter(age_ice >= min_year_vostok & age_ice <= max_year_vostok)

ggplot(vostok_temp_subset, aes(x = age_ice/1000, y = temperature)) +
  geom_line(color = "blue", size = 1) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(x = "thousand years before present", y = "Temperature variation [C]",
       title = "Temperature change recorded in Vostok ice core (128357 - 138193 years ago).",
       caption = "Data: Carbon Dioxide Information Analysis Center (CDIAC)") +
  theme_standard

```

Next, fit a linear model to your subset

```{r}

# fit linear regression
score_model <- lm(temperature ~ age_ice, data = vostok_temp_subset)

# view summary of results
summary(score_model)

```

`r msmbstyle::question(label = "ques:cch-21", text = "Identify two more ranges with increasing temperatures and determine the rate of change during that time period.")`

Now, let's take a look at past patterns of changes in atmospheric CO<sub>2</sub> over time.

```{r eval=FALSE}

# define url
url <- "http://cdiac.ess-dive.lbl.gov/ftp/trends/CO<sub>2</sub>/vostok.icecore.CO<sub>2</sub>"

# download file
download.file(url, "data/vostok_ice.txt")

# load dataset
vostok_ice <- read_delim("data/vostok_ice.txt", delim = "\t",
                     skip = 21,
                     col_names = c("depth", "age_ice", "age_air", "CO<sub>2</sub>"))

```

`r msmbstyle::question(label = "ques:cch-22", text = "Apply what you have learned identifying past period of rapid temperature increase and calculating the rate of change and do the same for the CO<sub>2</sub> data set.")`

`r msmbstyle::question(label = "ques:cch-23", text = "Use your results to compare current changes (approx. 200 years) in atmospheric and CO2 concentrations and global temperature to pre-historic changes.")`

`r msmbstyle::question(label = "ques:cch-24", text = "Interpret your results to assess whether recent changes in temperature are due to natural vs anthropogenic factors. For your answer consider both arguments that attribute the change to anthropogenic factors and natural fluctuations.")`


## Acknowledgments

This tutorial is based on the EDDIE Climate Change Module^[O'Reilly, C.M., D.C. Richardson, and R.D. Gougis. 15 March 2017. Project EDDIE: Climate Change. Project EDDIE Module 8, Version 1.]







