---
title: "Data Science for the Natural Environment (FA21)"
subtitle: "BI449 Data Science for the Natural Environment"
author: "Shannon J. O'Leary"
date: "`r Sys.Date()`"
knit: "bookdown::preview_chapter"
output:
  msmbstyle::msmb_html_book:
    highlight: tango
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    margin_references: FALSE
    css: msmb.css
bibliography: labmanual.bib
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}

library(msmbstyle)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      cache.extra = packageVersion('msmbstyle'),
                      warning = FALSE,
                      message = FALSE)


options(htmltools.dir.version = FALSE)

source("scr/ggplot.R")

theme_set(theme_standard)

```

# Remote Sensing

**Learning Objectives**

After completing this tutorial you should be able to

* understand the importance of long-term monitoring stations and baseline data
* understand how spatial and temporal patterns affect ecological processes
* describe what raster data sets are and how they encode spatial information
* read spatial raster data in R, making simple maps, and extracting information for non-spatial statistical tests.
* test whether plant growth (greenness & height) is drive more by elevation, slope, or aspect.
* consider large scale patterns of relationships between topography and vegetation

Download the [06_RemoteSensing](https://drive.google.com/drive/folders/1r2B4wa0hRGxK8aoNaoeQhULn3j2cFZNA?usp=sharing) project folder. Once you have downloaded it, unzip the project directory into your `BI449` directory.

Your first step is creating a new R project in your project folder. To do this open `Rstudio` then use the drop down menu in the top right corner to select `New Project` and from there `New Project in an existing directory`. From there, navigate to and select the project folder create an R project. Remember, your `Rproj` sets the working directory so you need to make sure that it is in your project folder.

Next, open a new `Rmd` file using `File > New File > R Markdown` or the drop down menu of the green plus button below `File` and save it in your Remote Sensing project directory. Use the setup chunk to set your global settings for the document using the options you've learned about in the last few projects. Remember, you should always check your `html` document after it knits to make sure that everything has converted as expected, including your figures, bullet points etc. We are going to need some specific packages for mapping; if you have not already you will need to install the following^[Remember, you only need to do this once, so you should type these functions directly into the console rather than have a code chunk which will try to run every time you knit the document.]

```{r evaluate=FALSE}

install.packages("rnaturalearthhires", 
                 repos = "http://packages.ropensci.org", 
                 type = "source")

install.packages("ggspatial")

```

We will also need the package `sf`,`rgdal`, `raster`, and `corrplot`. If you are on a Mac and it asks you to install from source, say no.

Let's load our packages so we can get started.

```{r}

# load libraries
library(tidyverse)
library(janitor)
library(skimr)

library(rgdal)
library(raster)
library(ggplot2)
library(viridisLite)
library(corrplot)


# won't need this set of libraries 
# unless you want to plot the final figure yourself
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthhires)
library(rnaturalearthdata)

# turn of sci notation
options(scipen=999)

```

## NEON & the power of longterm data sets

`r msmbstyle::question_begin(label = "ques:rem-")`

Watch this [short introduction](https://www.youtube.com/watch?v=38JQXPubMoA) to the National Ecological Observatory Network (NEON) and read this [overview of the field stations and domains](https://www.neonscience.org/field-sites/about-field-sites) withing the network of monitoring stations.

Briefly outline what NEON is, how it is designed, what data is being measured, and what the central goal is and how the design supports this mission.

Discuss the value of long-term monitoring stations and standardized data gathering.

`r msmbstyle::question_end()`


## Plot maps of topography and plants

The data set we are going to start out with is from the Sierra Nevada mountains in California from a region called [Soaproot Saddle](https://www.neonscience.org/field-sites/soap). 

We are going to load three raster data sets.

`r msmbstyle::question_begin(label = "ques:rem-")`

Briefly describe how a raster data set encodes information.

`r msmbstyle::question_end()`

Load **Digital Terrain Model** data (DTM).

```{r}

dtm <- raster("data/NEON_D17_SOAP_DP3_298000_4100000_DTM.tif")

```

`r msmbstyle::question_begin(label = "ques:rem-")`

Briefly describe what data the terrain model raster data set contains.

`r msmbstyle::question_end()`

Load **Digital Surface Model**, LiDAR first returns, topography + vegetation

```{r}

dsm <- raster("data/NEON_D17_SOAP_DP3_298000_4100000_DSM.tif")

```

`r msmbstyle::question_begin(label = "ques:rem-")`

Briefly describe what data the surface model raster data set contains.

`r msmbstyle::question_end()`

Load **Normalized Difference vegetation index** (NDVI)

```{r}

ndvi <- raster("data/NEON_D17_SOAP_DP3_298000_4100000_NDVI.tif")

```

`r msmbstyle::question_begin(label = "ques:rem-")`

Describe what NDVI measures and how it can be used to assess vegetation.

`r msmbstyle::question_end()`

We can calculate the vegetation height as DSM - DTM.

```{r}

veg_ht <- dsm - dtm

```

Next, let's stack the data so we can plot it all in the same plot

```{r}

# create raster stack
all_data <- stack(dtm, dsm, ndvi, veg_ht)

# rename raster layers
names(all_data) <- c("DTM", "DSM", "NDVI", "Vegetation.Height")

```

Now we can create maps for each of our raster files^[this may take a second to plot ... you are processing a bunch of data!].

```{r}

plot(all_data, 
     col = viridis(255))

```

We already have information on elevation, but let's calculate slope and aspect as additional metrics to describe the topology.

We will use degrees as a measure for slope.

```{r}

slope <- terrain(dtm, opt = "slope", unit = "degrees", neighbors = 8)

```

For aspect we will calculate "northness" (cosine of aspect) which will read in radians^[Radian is the SI unit for measuring angles; it is a dimensionless value.].

```{r}

# calculate aspect
aspect <- terrain(dtm, opt = "aspect", unit = "radians", neighbors = 8)

# calculate northness
north <- cos(aspect)

```

Let's add these layers to our raster stack to make it easy to plot everything in a single plot.

```{r}

all_data$slope <- slope

all_data$north <- north

```

Now we can plot all six maps.

```{r}

plot(all_data, 
     col = viridis(255))

```

Let's consider what correlations we would expect to find based on these six maps.

`r msmbstyle::question_begin(label = "ques:rem-")`

Argue whether overall you think that vegetation patterns are influenced by topography based on these maps and list anything noteworthy you have observed about these maps.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Argue which topographic metrics you think should have a strong association with NDVI and describe what you would expect that relationship to look like.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Argue which topographic metric you think should have a strong association with vegetation height and describe what you would expect that relationship to look like.

`r msmbstyle::question_end()`


## Explore relationship between topographic variable and vegetation patterns

First, lets convert our raster layers into data frames for easier use. We'll start by creating an empty data frame with as many rows as there are cells in our `raster` objects. Then we can use the function `extract()` the entire extent of the raster layer, i.e. we are pulling out all the information for each pixel and putting it into a set of columns[^3]. The `extent` argument, tells R to pull out all the information in the `raster` object^[Recall, each column of a dataframe is a vector].

```{r}

# create and empty dataframe for our data set
df <- as.data.frame(matrix(NA, nrow = ncell(dtm), ncol=0))

# extract vegetation height
df$veg_ht <- extract(veg_ht, extent(veg_ht))

# extract ndvi
df$ndvi <- extract(ndvi, extent(ndvi))

# extract dtm
df$dtm  <- extract(dtm, extent(dtm))

# extract slope
df$slope  <- extract(slope, extent(slope))

# extract aspect
df$north <- extract(north, extent(north))

head(df) %>%
  kable()

```

Note, that we no longer have spatial information, however, values in the same row do correspond to the same pixel location in our raster object.

Because we have so much data, we will take a 1% subset of the data to decrease the computation power and time needed. Because our data set is so large (1 Million rows!), we would expect that a random subset is representative of the relationships as a whole.

`dplyr::slice_sample()` can be used to specify the proportion of rows that you would like to retain. The function will return a random subsample.

```{r}

# set seed for reproducibility
set.seed(42)

df_sub <- df %>%
  slice_sample(prop = 0.01)

```

Let's take a look at the relationship between vegetation variables and topographic variables. To do this efficiently using the `tidyverse` principles, need to pivot our data set. Ultimately, we want to have a data set with one column with our topography parameters, one with the topography measurements, one with vegetation variables and one with measurements.

```{r}

df_plot <- df_sub %>%
  pivot_longer(names_to = "topog_param", values_to = "topog_meas", 3:5) %>%
  pivot_longer(names_to = "veget_param", values_to = "veget_meas", 1:2)

ggplot(df_plot, aes(x = topog_meas, y = veget_meas)) +
  geom_point(alpha = 0.25, size = .75) +
  facet_grid(veget_param ~ topog_param, scales = "free") +
  geom_smooth(method = "lm")

```

`r msmbstyle::question_begin(label = "ques:rem-")`

Use the scatterplots to make predictions about statistical relationships. Consider whether any of these plots look like they are visualizing strong relationships.

Argue whether you think larger data sets make it easier or harder to identify distinct relationships "by eye".

`r msmbstyle::question_end()`

Let's calculate the correlations and determine whether or not they are significant.

We can calculate a correlation using Pearson's R and test for significance using the function `cor.test()`

```{r}

cor.test(df_sub$dtm, df_sub$veg.ht)

```

`r msmbstyle::question_begin(label = "ques:rem-")`

Calculate the remaining correlations and record them in the table below.

`r msmbstyle::question_end()`

Let's record all of these values in a table (I have added the first one for you).

|               |         | DTM | Slope | Northness |
|---------------|---------|-----|-------|-----------|
| NDVI          | R       |     |       |           |
|               | p-value |     |       |           |
| Veget. height | R       |-0.156  |       |           |
|               | p-value |< 0.001 |       |           |
|

`r msmbstyle::question_begin(label = "ques:rem-")`

Describe whether or not these relationships conform to your predictions when you looked at the maps and when you looked at the scatterplots and discuss what could be causing differences from your expectations.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Assess whether or not all of your relationships are statistically significant. Discuss the importance of p-values for large data sets.

`r msmbstyle::question_end()`


## Comparison of relationship of vegetation and topography across the US

Let's take a look at the different [NEON sites](https://www.neonscience.org/field-sites/explore-field-sites) across the US and choose locations from the contiguous 48.

`r msmbstyle::question_begin(label = "ques:rem-")`

List the sites that were chosen and describe the sampling design behind using this specific set of locations.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Briefly describe what patterns you expect to see, consider e.g. whether you think that you will observe the same relationship across all locations. Look up the sites you have chosen to make informed predictions.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

For your assigned sites go through the following process:

1.  read in the DTM, DSM, and NDVI `*tif` files as `raster` objects.
2.  Calculate vegetation heigth, slope, and northness and create a raster stack to plot the six panel plot.
3.  Based on your plots predict which relationship you think will have a stronger association with vegetation height, determine if vegetation patterns overall appear to be influenced by topography, and note any other observations for this set of maps on their own and in relation to the California set.
4.  conver your raster table to a data.frame and take a 1% subset.
5.  plot the correlations between vegetation and topographic variables.
6.  calculate the correlations and record them in a table.

Share your correlations in the #hwassignments channel on slack

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Compare the overall patterns of correlation across the United States. Identify patterns and determine if the relate to geographic variables discussed in class. Identify mechanisms that could explain these observations.

`r msmbstyle::question_end()`

`r msmbstyle::question_begin(label = "ques:rem-")`

Consider differences in scale and describe how our results might differ if we had lower (bigger pixels) or higher (smaller pixels) resolution or we consider larger spatial extents for each location. Consider how temporal change (e.g. droughts) might impact the observed patterns.

`r msmbstyle::question_end()`
