---
title: "Data Science for the Natural Environment (FA21)"
subtitle: "BI449 Data Science for the Natural Environment"
author: "Shannon J. O'Leary"
date: "`r Sys.Date()`"
knit: "bookdown::preview_chapter"
output:
  msmbstyle::msmb_html_book:
    highlight: tango
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    margin_references: FALSE
    css: msmb.css
bibliography: labmanual.bib
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}

library(msmbstyle)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      cache.extra = packageVersion('msmbstyle'),
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE)


options(htmltools.dir.version = FALSE)

source("scr/ggplot.R")

theme_set(theme_standard)

```

# Predictive modeling for air pollution

**Learning Objectives**

After completing this tutorial you should be able to

* 

Let's load our packages so we can get started.

```{r}

# load libraries for data import & wrangling
library(magrittr)
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(knitr)


# load libraries for data visualization & mapping
library(ggplot2)
library(patchwork)
library(sf)
library(osmdata)


# turn of sci notation
options(scipen=999)

```


Read in air pollution data

```{r}


pollution <- read_delim("data/airpollution_raw.txt", delim = "\t") %>%
  clean_names() %>%
  filter(date > 0 & time > 0) %>%
  separate(time, into = c("hr", "mi", "sc"), sep = c(2,4), remove = FALSE, convert = TRUE) %>%
  mutate(date = ymd(date),
         weekday = weekdays(date),
         year_month = format_ISO8601(date, precision = "ym"),
         week = week(date),
         dy = day(date),
         yr = year(date),
         mt = month(date),
         datetime_UTC = make_datetime(year = yr, month = mt, day = dy,
                                  hour = hr, min = mi, sec = sc),
         datetime_EST = with_tz(datetime_UTC, tzone = "EST")) %>%
  filter(yr > 2020)

glimpse(pollution)

```

Check out variables

```{r}

skim(pollution)

```

Find distribution of PM2.5 values

```{r}

pollution %>%
  select(pm2_5) %>%
  summary()

quantile(pollution$pm2_5, probs = c(0.05, 0.25, 0.5, 0.75, 0.95, 0.99))

```

Plot distributions

```{r}

ggplot(pollution, aes(x = pm2_5)) +
  geom_histogram()

```

Check out the high values; all from certain outings or distributed across?

WHO

* PM2.5 > 5ug/m3

AQI: 

* Good: 0-12
* Moderate: 12-35
* Unhealthy for sensitive individuals: 35-55
* Unhealthy: 55-150
* Very Unhealthy: 150-250
* Hazardous: > 250

Pull out Unhealthy or higher

```{r}

pollution %>%
  filter(pm2_5 > 55) %>%
  group_by(id) %>%
  count() %>%
  arrange(desc(n))

```

Pull out very unhealthy or higher

```{r}

pollution %>%
  filter(pm2_5 > 150) %>%
  group_by(id) %>%
  count() %>%
  arrange(desc(n))

```


Pull out hazardous

```{r}

pollution %>%
  filter(pm2_5 > 250) %>%
  group_by(id) %>%
  count() %>%
  arrange(desc(n))

```

Check out that one recording

```{r}

hazard <- pollution %>%
  filter(id == "E390051.CSV")

ggplot(hazard, aes(x = datetime_EST, y = pm2_5)) +
  geom_line() +
  scale_y_log10()

```

One location far from Manchester

```{r}

mass <- pollution %>%
  filter(id == "CE30009.CSV")

```



Remove from data set

```{r}

pollution %<>%
  filter(!id %in% c("E390051.CSV", "CE30009.CSV"))

```

Compare by week day

```{r}

pollution %<>%
  mutate(weekday = ordered(weekday, levels = c("Monday",
                                               "Tuesday",
                                               "Wednesday",
                                               "Thursday",
                                               "Friday",
                                               "Saturday",
                                               "Sunday")))

ggplot(pollution, aes(x = weekday, y = pm2_5)) +
  geom_boxplot() +
  geom_hline(yintercept = c(12, 35, 55, 150, 250)) +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(pollution, aes(x = weekday, y = pm2_5)) +
  geom_boxplot() +
  geom_hline(yintercept = c(12, 35, 55, 150, 250)) +
  scale_y_log10() +
  coord_flip()

```

Compare by hour

```{r}

pollution %<>%
  mutate(hr_EST = factor(hour(datetime_EST))) %>%
  mutate(AQI = case_when(pm2_5 <= 12 ~ "good",
                                pm2_5 > 12 & pm2_5 <= 35 ~ "moderate",
                                pm2_5 > 35 & pm2_5 <= 55 ~ "unhealth sens",
                                pm2_5 > 55 & pm2_5 <= 150 ~ "unhealthy",
                                pm2_5 > 150 & pm2_5 <= 250 ~ "very unhealth",
                                pm2_5 > 250 ~ "hazardous"))

ggplot(pollution, aes(x = hr_EST, y = pm2_5)) +
  geom_boxplot() +
  facet_grid(AQI ~ ., scales = "free") +
  scale_y_log10()

```

Compare mean per hour for days with measurement

```{r}

hourly <- pollution %>%
  group_by(date, hr_EST) %>%
  summarize(mean_hr = mean(pm2_5),
            max_hr = max(pm2_5),
            hr_05 = quantile(pm2_5, probs = .05),
            hr_95 = quantile(pm2_5, probs = .95)) %>%
  pivot_longer(names_to = "stat", values_to = "pm2_5", 3:6)

ggplot(hourly, aes(x = date, y = hr_EST, fill = log10(pm2_5))) +
  geom_tile() +
  facet_grid(stat ~ .) +
  scale_fill_viridis_c()

ggplot(hourly, aes(x = date, y = hr_EST, fill = mean_hr)) +
  geom_tile() +
  scale_fill_viridis_c()

```

Look at spatial component

```{r}

poll_spat <- pollution %>%
  filter(latitude > 0 & longitude > 0)

```

Ge map info

```{r}

# available categories for features
available_tags("highway")
available_tags("waterway")
available_tags("railway")
available_tags("water")

# get geographic location of a city
getbb("Manchester New Hampshire")

# pull specified set of street categories & convert to sf
prim_roads <- getbb("Manchester NH") %>%
  opq() %>%
  add_osm_feature(key = "highway",
                  value = c("motorway", "primary", "motorway_link", "primary_link")) %>% 
  osmdata_sf()

second_roads <- getbb("Manchester NH") %>%
  opq() %>%
  add_osm_feature(key = "highway",
                  value = c("secondary", "tertiary", "secondary_link", "tertiary_link")) %>% # small roads
  osmdata_sf()

small_roads <- getbb("Manchester NH") %>%
  opq() %>%
  add_osm_feature(key = "highway",
                  value = c("residential", "living_street",
                            "unclassified",
                            "service", "footway")) %>% # small roads
  osmdata_sf()

rivers <- getbb("Manchester NH") %>%
  opq() %>%
  add_osm_feature(key = "water",
                  value = c("canal", "river", "stream", "tidal_channel", "basin", "lake", "reservoir")) %>% 
  osmdata_sf()

haz_spatial <- hazard %>%
  filter(latitude > 0 & longitude > 0) %>%
  mutate(longitude = -1*longitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

tmp <- hazard %>%
  filter(latitude > 0 & longitude > 0)

unhealthy <- poll_spat %>%
  filter(pm2_5 > 35) %>%
mutate(longitude = -1*longitude) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)


x_min <- -1*min(tmp$longitude) - .1
x_max <- -1*max(tmp$longitude) + .1

y_min <- min(tmp$latitude) - .1
y_max <- max(tmp$latitude) + .1

ggplot() +
  geom_sf(data = prim_roads$osm_lines,
          inherit.aes = FALSE,
          color = "black") +
  geom_sf(data = second_roads$osm_lines,
          inherit.aes = FALSE,
          color = "black") +
  # geom_sf(data = small_roads$osm_lines,
  #         inherit.aes = FALSE,
  #         color = "#666666") +
  # geom_sf(data = second_roads$osm_lines,
  #         inherit.aes = FALSE,
  #         color = "black",
  #         size = .3,
  #         alpha = .5) +
  # geom_sf(data = small_roads$osm_lines,
  #         inherit.aes = FALSE,
  #         color = "#666666",
  #         size = .2,
  #         alpha = .3) 
  geom_sf(data = unhealthy, shape = 21, size = 2, alpha = .75, aes(fill = pm2_5)) +
  scale_fill_viridis_c() +
  coord_sf(xlim = c(42.8, 43), 
           ylim = c(-71.55, -71.4),
           expand = FALSE)

```


```{r}

df_sf <- pollution %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

grid <- df_sf %>% 
  st_bbox() %>% 
  st_as_sfc() %>% 
  st_make_grid(cellsize = 0.05)

grid_sf <- grid %>%
  st_sf() %>%
  rownames_to_column("polygon_id") %>%
  mutate(polygon_id = as.integer(polygon_id))

pollut_grid <- poll_spat %>%
  mutate(polygon_id = st_intersects(df_sf, grid) %>% map_int(1))

plot <- pollut_grid %>%
  group_by(polygon_id) %>%
  summarize(mean_pm = mean(pm2_5),
            max_pm = max(pm2_5),
            pm_05 = quantile(pm2_5, probs = .05),
            pm_95 = quantile(pm2_5, probs = .95)) %>%
  left_join(grid_sf, by = "polygon_id")

ggplot() +
  geom_sf(data = plot, aes(fill = pm_95, geometry = geometry)) +
  scale_fill_viridis_c()


```

